A Bayesian approach to burial networks to infer social changes in northeastern Taiwan during the European colonization period

### 2021-01-11 analysis/paper/paper.Rmd

<!--fit in a bigger historical picture in a periclonial context -->
The involvement in long-distance exchange is often associated with changes in political-economic strategies to a network mode, where ambitious individuals are able to build their local power base and personal influence through the distribution of high-value goods [@Blanton1996; @Feinman2000; @Klehm2017]. Northeastern Taiwan was involved in a regional exchange in East Asia since the 14th century and long-distance trade during the European presence in the 17th century [@Chen2005; @Wang2007]. In 1626, the Spanish built their forts in Keelung and Tamsui, northern Taiwan, and later were defeated and taken over by the Dutch in 1642 [@Andrade2007]. Indigenous people in northeastern Taiwan first encountered colonial power in 1632 when they were attacked by the Spanish and later in 1647 by the Dutch, who also asked them to pay annual tributes in a flexible way [@Borao2009; @Kang2016]. The Dutch would give local Indigenous leaders symbolic items or trade goods to assert their political authority and consolidate their relationship with Indigenous societies.

<!-- archaeological specific case study -->
In this study, we explore the changes in social inequality in a pericolonial context at Kiwulan, an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the network of our model? (2) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

@@

<!--archaeological setting-->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD). The settlement experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

@@

@@

<!-- Network construction and specification -->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:ergm-terms-table) lists the configurations we used for the model specification for burials with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality that is represented by endogenous network effects, low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to test whether burials with similar attributes tend to form relationships. For example, age-homophily means people of the same age to have the same burial goods. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to control the preferential tendency of formation of relationship. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

```{r bergm-iteration}
burn_in <- 100
main_iter <- 1000
aux_iter <- 4000
```

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distributions for the priors are typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of the edge density parameter to low for both network models. For covariates based on burial attributes, such as age, sex, and our wealth index, we specified a vague prior that follows a normal distribution with mean at 0 and standard deviation at 3 (i.e. N(0, 3)) for both models. For physical distance between burials, we set a negative covariate effect (N(-1, 1)) to infer kinship-based proximity, e.g. stronger correlations for shorter distances. To evaluate our anthropological assumption about increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for social inequality, especially for transitivity (gwesp) and popularity (gwedegree). We set the priors to higher transitivity (N(2, 4)), lower popularity (N(-1, 1)), and higher covariate effect based on ritual activity (N(1, 5)) for network before European contact to indicate less social inequality. On the contrary, we set the priors for the network after the European arrival to lower transitivity (N(1, 2)), and higher popularity (N(1, 2)) to model an increased social inequality. This prior information derives from theory about horizontal hierarchies, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate mode at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set the same with N(-3, 3) for both pre- and post-European networks.

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
As an informal way to diagnose model convergence, we first evaluated three diagnostic visual summaries of our MCMC output, including density plots, trace plots, and autocorrelation plots [@Hamra2013], which we include in our Supplementary Online Materials. In general, we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation. Then we summarize output from our two models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. We set 10,000 network graphs simulated from the estimated posterior distribution in ERGMs. Our Bayesian GOF diagnostics summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution, based on our hypotheses, can reproduce networks with similar general structural features of the observed networks. We then compared the distribution of our observed networks, the networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network, such as the relative importance of age and sex.

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For the pre-European network simulations, two nodal covariates, age-homophily and ritual-homophily, have significant effects on the formation of relations between burials. The age homophily shows a negative effect that indicates burials of the same age tend to not share similar prestige goods, while the ritual-homophily demonstrates the positive effect that means burials with ritual pottery tend to form relations. Despite positive values for some covariates, such as sex-homophily and burial value-homophily, they do not show a significant tendency to form relations duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows positive effects but not significantly, indicating that physical proximity does not affect relations. For the endogenous network effects, transitivity (gwesp) present significant positive effect, while popularity (gwdegree) demonstrates negative effect. The positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. The negative popularity shows there is a tendency toward decentralization that reflects most burials have similar number of ties without any prominent ones. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, GWESP, GWDeg, and distance-dyadcov by phases. The pre-European group presents remarkedly larger values for the GWESP, but lower vlaues for the edges and GWDeg parameters comapred to the post-Eruoepan group. The distance parameters overlap between two groups."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
For the network after the arrival of Europeans the nodal covariates of age-homophily, sex-homophily, ritual-homophily, and the dyadic covariate of physical distance all show positive effects, while the burial value-homophily has a negative effect. However, there are no significant effects for these nodal and dyadic covariates. We also found that the burials in the same wealth level tend not to form relations. Similar to the pre-European network, the endogenous network variable, transitivity (gwesp) demonstrates a significant positive effect. In contrast to the pre-European network, the popularity (gwdegree) shows a significant positive effect. This means there is a tendency toward centralization that reflects a limited number of burials have many more ties than others. This implies that a few central individuals may have better access to trade goods and developed personal networks through wealth accumulation. In general, the post-European network has a smaller transitivity effect and a positive popularity effect than the pre-European network . This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence. Both posterior estimates present symmetric distributions, with the posterior means located close to posterior medians (Table XX).

```{r vertex-bootstrap}
boostrap <- 1000
```

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used vertex bootstrap to cross-validate the results based on Bayesian network analysis. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) that is useful for quantifying standard errors and estimating sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks by examining their confidence intervals for the network population. Networks with unoverlapped intervals means there is a significant difference. We set the resampling size to 1,000 networks to compute endogenous network statistics, including density, transitivity, and popularity for our two networks. The result shows a significant difference in network popularity between the two networks, which is consistent with our finding of negative popularity in the pre-European period and positive popularity in the post-European period using the Bayesian approach. For density and transitivity, the 95% confidence intervals from two networks overlapped with each other that indicates no significant difference. This is also consistent with our results of Bayesian modeling, where both networks present similar positive or negative effects some degree of difference.

Using the same vertex bootstrap procedure, we can also test the difference of network populations with a t-test on density, transitivity, and popularity. The results show that there are significant differences in all network statistics. Moreover, we explored the sample size effect by removing nodes at certain percents (5-40%) for both networks and compared their confidence intervals. The results demonstrate a consistent difference between the two networks that means they are robust to the different sizes. In general, those statistical methods support the presence of anthropologically meaningful differences in the pre-and post-European networks. The procedure of cross-validation ensures our results can provide robust comparison between two networks without sample size effects.

```{r Bayesian-gof}
# the grey lines mark the 95% intervals. X-axis means the proportion of nodes
```

<!--assessment of MCMC analysis and BGOF--> 
Bayesian goodness of fit (GOF) diagnostics plots show that both models fit the observed networks very well for the minimum geodesic distance distribution (i.e. the number of edges between node pairs in a shortest path [@Hunter2008]) and the degree distribution. For edgewise shared partner (gwesp) distribution, despite some observations falling outside the 95% interval, the fit is generally good with most observations within it. This also suggests that our observed network statistics do not follow a normal distribution, especially for the network after the arrival of the Europeans that presents a bimodal pattern. We compared the first three distribution moments of each observed distribution and their corresponding simulated distributions, represented by means. Figure \@ref(fig:figure-distribution-moments) shows that modeled values from the pre-European network are closer to the observed values overall, compared to the values from the post-European network. For the variance of degree, the values from the modeled post-European distribution are closer to the observed distribution. In general, this demonstrates a better fit for the pre-European network. However, it should be noted that the differences between the statistics of two models are relatively small, which still allows direct comparison of two networks.

```{r figure-distribution-moments, fig.cap = "Comparison of statistics calculated on the observed and simulated GOF distributions for models between phase groups "}
knitr::include_graphics(here::here("analysis", "figures", "006-distribution-moments.png"))
```

@@

<!--archaeological implication based on the results-->
The covariants for both networks, before and after the European presence, provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:xxx)). Our covariants were age, sex, ritual pottery, and a wealth index based on the value of grave goods. The pre-European network shows that burials with pottery as ritual vessels tend to be tied. On the other hand, in the post-European network ties tend to be between burials within the same level of wealth. Moreover, in the post-European network, burials with ritual vessels tend not to be tied. This illustrates that different social and economic mechanisms determined relations between burials for the two periods. Before European arrival, ritual behaviors seem to have been important status indicators, relative to the use of prestige goods. After European presence was established, status indicators shifted from ritual to wealth differences. Physical distance between burials did not correlate to network formation for either period. This means geographical proximity, implying some kinship association, is not a factor for sharing similar goods between burials. This may suggest that the use of burial goods is more individual-oriented instead of kinship-based. In addition, demographic information, such as age and sex, do not show any correlations with network formation for either period. This may be due to the unavailability of age and sex values for many burials in our sample.

>>>>

<!--fit in a bigger historical picture in a periclonial context -->
The involvement in long-distance exchange is often associated with changes in political-economic strategies to a network mode, where ambitious individuals are able to build their local power base and personal influence through the distribution of high-value goods [@Blanton1996; @Feinman2000; @Klehm2017]. Northeastern Taiwan was involved in a regional exchange in East Asia since the 14th century and long-distance trade during the European presence in the 17th century [@Chen2005; @Wang2007]. In 1626, the Spanish built their forts in Heping dao and Tamsui (Figure \@ref(fig:figure-KWL-map)), northern Taiwan, and later were defeated and taken over by the Dutch in 1642 [@Andrade2007]. Indigenous people in northeastern Taiwan first encountered colonial power in 1632 when they were attacked by the Spanish and later in 1647 by the Dutch, who also asked them to pay annual tributes in a flexible way [@Borao2009; @Kang2016]. The Dutch would give l
ocal Indigenous leaders symbolic items or trade goods to assert their political authority and consolidate their relationship with Indigenous societies.

<!-- archaeological specific case study -->
In this study, we explore the changes in social inequality in a pericolonial context at Kiwulan (Figure \@ref(fig:figure-KWL-map)), an Iron Age site in northeastern Taiwan, which covers the time before the European arrival, the presence of the Europeans, including the Spanish and the Dutch, in the 17th century, and finally the Chinese in the 19th century [@Chen2007]. We assume that the social changes at Kiwulan would be supported by chronological differences in the structure of burial networks. We examine burial networks before and after the foreign contacts, and test our hypothesis that increased social inequality can be observed in the network after the European arrival. We ask: (1) does the observed burial data after the start of European presence resemble the network of our model? (2) did European colonial activities in 17th century Taiwan result in an increased social inequality in an Indigenous society in ways that can be detected by burials networks? (3) what are the major variables affecting or forming the higher degree of popularity (e.g. a few nodes have more relationships with other nodes) in the European contact period that might hint at social heterogeneity during this period? By answering these questions, this study helps to expand the use of statistical modeling for burials in understanding the indirect effects of a colonial presence on Indigenous groups. This is important to understand the reactions of Kiwulan residents in the context of indirect impacts of European colonization.

```{r, figure-KWL-map, fig.cap = "The location of Kiwulan and other places with the European forts in northern Taiwan mentioned in the text (map modified from Wang and Marwick, 2020). Map data from naturalearthdata.com."}

knitr::include_graphics(here::here("analysis", "figures", "000-KWL-map.png"))
```

@@

<!--archaeological setting-->
We analyzed burial data collected from the published excavation reports, and the original fieldwork notes for the upper component of Kiwulan, an Iron Age settlement (1350-1850 AD). The settlement experienced European colonial impacts in the early 17th century and a large wave of Chinese immigrants in the 19th century [@Chen2007]. Excavations revealed abundant pottery sherds, imported ceramics and stonewares, wooden artifacts, stone tools, metal artifacts, imported glass beads and agates beads, and pipes [@Chen2007]. In addition to these artifacts, 90 burials, hundreds of middens, storage pits, and postholes with _in-situ_ posts were also excavated. The burials are mostly located in the middle section of the excavation area, which is the largest open area at Kiwulan that provides continuous stratigraphic sections suitable for temporal comparison (Figure \@ref(fig:figure-burial-map)). Burials are oriented in an east-west direction on the north side of the residential area, indicated by post-holes and `in-situ` wooden posts, indicating a well organized spatial arrangement. Previous studies report an uneven distribution of prestige goods across burials without agreement about whether this uneven distribution hints at vertical social differences. For example, @Cheng2008 interpreted the unequal distribution of glass beads, especially the gold-foil beads between burials, as evidence for hierarchy, indicating a stratified society. However, @Hsieh2012 suggested a relatively egalitarian structure based on comparative analysis of the frequencies of all burial goods. She found that the burials with high value burial goods were usually associated with elders, which might indicate achieved, rather than inherited, status. One important limitation of these previous studies is that they did not use analytical units suitable for comparing pre-European social organization with post-European social organization. Here, we adopt a new chronological framework for the burials to test if network configurations differ from the pre-European period to the post-European period. The discord over the discussion of Kiwulan social organization could be associated with chronological differences that has not yet been well studied.

@@

<!-- Network construction and specification -->
We used Bayesian inference on exponential random graph models (ERGMs) to quantify the relations among burials and test our hypothesis of social change using the R programming language [@Rlanguage2019] with the bergm package [@Caimo2014], which is built upon the statnet package for model specification and simulation procedures [@Hunter2008; @Morris2008; @Handcock2008]. Table \@ref(tab:table-ergm-terms) lists the configurations we used for the model specification for burials with the corresponding archaeological evidence. Every parameter in an ERGM has an associated algorithm for computing the probability of observing relations in terms of grave goods between two burials. Based on our hypothesis, we model a network with increased social inequality that is represented by endogenous network effects, low transitivity and high popularity. We include burial-specific attributes as covariate effects for homophily, such as age, sex, ritual activity, and the degree of wealth, to test whether burials with similar attributes tend to form relationships. For example, age-homophily means people of the same age to have the same burial goods. We also take into account the physical distance between burials as an indicator of a kinship-based relation since the deceased from the same family tend to be buried nearby [@LiandWu2006]. Our model could reveal the emergence of social inequality via the presence of a few individuals as network centers, having more relations with others. This would be indicated by high popularity or degree values in the network statistics, with covariates to control the preferential tendency of formation of relationship. According to our hypothesis, the burial evidence from after the European arrival will show higher popularity compared with the burial evidence from before European contact.

@@

```{r bergm-iteration}
burn_in <- 100
main_iter <- 1000
aux_iter <- 4000
bgof_si <-10000
```

@@

<!-- add text about prior setting and anthropological assumptions of our prior values --> 
Normal distributions for the priors are typical in network analysis studies that assume networks to have low density and high transitivity, as are commonly found in the real world [@Caimo2017]. Thus, we specified the prior of the edge density parameter to low for both network models. For covariates based on burial attributes, such as age, sex, and our wealth index, we specified a vague prior that follows a normal distribution with mean at 0 and standard deviation at 5 (i.e. N(0, 5)) for both models. For physical distance between burials, we also set a vague prior effect (N(0, 1)) to explore whether there is kinship-based proximity, e.g. stronger correlations for shorter distances. To evaluate our anthropological assumption about increased social inequality over time, we incorporated different prior information for the network variables that are meaningful for social inequality, especially for transitivity (gwesp) and popularity (gwedegree). We set the priors to higher transitivity (N(2, 2)), lower popularity (N(-2, 3)), and higher covariate effect based on ritual activity (N(1, 5)) for network before European contact to indicate less social inequality and stress ritual element shared in corporate groups. On the contrary, we set the priors for the network after the European arrival to lower transitivity (N(1, 3)), higher popularity (N(2, 3)), and higher covariate effect of burial values (N(2, 3)) to model an increased social inequality. This prior information derives from theory about horizontal hierarchies, which can be viewed as a spectrum that illustrates an increasing social inequality from a corporate mode at one end toward a network mode at the other end [@Drennan2010; @Feinman2000]. The priors for density were also set the same with N(-3, 3) for both pre- and post-European networks based on a common observation of network density.

<!-- assessment of MCMC and Goodness-of-fit in Bayesian-->
As an informal way to diagnose model convergence, we first evaluated three diagnostic visual summaries of our MCMC output, including density plots, trace plots, and autocorrelation plots [@Hamra2013], which we include in our Supplementary Online Materials. In general, we can see the informative prior contributed to a better convergence of the MCMC according to the diagnostic plots showing more stationary distributions and decreased autocorrelation. Then we summarize output from our two models by goodness-of-fit (GOF) diagnostics in the Bayesian framework, where the observed network is compared with the set of networks simulated from the estimated posterior distributions of the parameters of each model [@Caimo2011; @Caimo2017]. We set `r bgof_si` network graphs simulated from the estimated posterior distribution in ERGMs. Our Bayesian GOF diagnostics summarized three distributions, including degree, minimum geodesic distance, and edgewise shared partner distributions. This provides a statistical approach to check how well the estimated posterior parameter distribution, based on our hypotheses, can reproduce networks with similar general structural features of the observed networks. We then compared the distribution of our observed networks, the networks before and after the arrival of Europeans, with the distribution of our hypothesized models. We expect to see the models fit with our hypotheses, indicative of a increased social inequality after the foreign contact. In addition, the covariates can give some more clues for the underlying mechanisms for the formations of each network, such as the relative importance of age and sex.

@@

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For the pre-European network simulations, the nodal covariates, ritual-homophily, have significant effects on the formation of relations between burials. The positive effect demostrates that burials with ritual pottery tend to form relations. Despite positive values for some covariates, such as age-homophily, sex-homophily and burial value-homophily, they do not show a significant tendency to form relations duo to value of zero in the confidence intervals. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not affect the formation of relations. For the endogenous network effects, transitivity (gwesp) present significant positive effect, while popularity (gwdegree) demonstrates negative effect. The high positive value for transitivity suggests a tendency of burials to be clustered in closed transitive structures, indicative of the presence of multiple corporate groups sharing burial goods in common. The negative popularity shows there is a tendency toward decentralization that reflects most burials have similar number of ties without any prominent ones. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

```{r figure-posterior-distribution, fig.cap = "Posterior density estimates for the parameters associated with edges, gwesp, gwdegree, and distance-dyadcov by phases. The pre-European group presents remarkedly larger values for the gwesp, but lower vlaues for the edges and gwdegree parameters comapred to the post-Eruoepan group. The distance parameters overlap between two groups."}
knitr::include_graphics(here::here("analysis", "figures", "006-posterior-distribution.png"))
```

<!--posterior stats for for post-E and their comparison-->
For the network after the arrival of Europeans, the nodal covariates of burial value-homophily show positive effects, while the ritual-homophily and sex-homophily have a negative effect. Despite the effect is not strong, it could indicate the burials in the same wealth level tend to form relations. On the contrary, the same sex burials and burial with ritual pottery in common tend not to form relations. There are no significant effects for age-homophily and the dyadic covariate, physical distance. Similar to the pre-European network, the endogenous network variable, transitivity (gwesp) demonstrates a significant positive effect but much lower than the effect of pre-European network. In contrast, the popularity (gwdegree) shows a significantly higher positive effect than the effect of pre-European network. This means there is a tendency toward centralization that reflects a limited number of burials have many more ties than others. This implies that a few central individuals may have better access to trade goods and developed personal networks through wealth accumulation. In general, the post-European network has a smaller transitivity effect and a positive popularity effect than the pre-European network. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence. Both posterior estimates present symmetric distributions, with the posterior means close to posterior medians (Table \@ref(tab:table-posterior-stats)).

```{r bootstrap-CI-test, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("007-bootstrap-CI.R"))}
suppressWarnings(source(here::here("analysis", "paper", "007-bootstrap-CI.R")))
```

<!-- comparison between two networks using bootstrap --> 
One key difference between the pre-European and post-European networks is their size, with `r pre_E` burials compared to `r post_E` burials. To understand the robustness of comparison between two networks, we used vertex bootstrap to cross-validate the results based on Bayesian network analysis. Vertex bootstrap is a non-parametric method that conducts resampling for all vertices (i.e. node) that is useful for quantifying standard errors and estimating sampling variability in the network statistics of interest [@Chen2019; @Snijders1999; @Rroberts2021]. This enables the evaluation of uncertainty for networks and tests the difference between multiple networks by examining their confidence intervals for the network population. Networks with unoverlapped intervals means there is a significant difference. We computed and compared endogenous network statistics, including density, popularity, and transitivity for our two networks. Figure \@ref(fig:figure-bootstrap-CI) shows a significant difference in observed network popularity between the two networks, which is consistent with our finding of negative popularity in the pre-European period and positive popularity in the post-European period using the Bayesian approach. For density and transitivity, the 95% confidence intervals from two networks overlapped with each other that indicates no significant difference. This is also consistent with our results of Bayesian modeling, where both networks present similar positive or negative effects with smaller degree of difference.

```{r figure-bootstrap-CI, fig.cap = "The 95% bootstrap confidence intervals for the observed networks, indicated by 0 nodes removal, and networks with different number of node removal. The number of bootstrap samples is 1000."}
knitr::include_graphics(here::here("analysis", "figures", "007-bootstrap-CI.png"))
```

```{r bootstrap-t-test, include=FALSE, cache=TRUE, cache.extra=tools::md5sum(here::here("008-bootstrap-t-test.R"))}
suppressWarnings(source(here::here("analysis", "paper", "008-bootstrap-t-test.R")))
pval_den <- signif(density_b_test$p.value, 4)
pval_trans <- signif(trans_b_test$p.value, 4)
pval_star <- signif(star_b_test$p.value, 4)
```

The vertex bootstrap procedure with a t-test on the difference of network populations also shows that there are significant differences in network density(p = $`r pval_den`$), transitivity (p = $`r pval_trans`$), and popularity (p = $`r pval_star`$). Moreover, we explored the sample size effect by removing nodes at certain percents (5-40%) for both networks and compared their confidence intervals. The results (Figure \@ref(fig:figure-bootstrap-CI)) demonstrate a consistent difference between the two networks within 30 % node removals that suggest they are robust to the different sizes. In general, those statistical methods support the presence of anthropologically meaningful differences in the pre-and post-European networks. The procedure of cross-validation ensures our results can provide robust comparison between two networks without sample size effects.

```{r figure-bgof-two-phases, fig.cap = "Goodness-of-fit diagnostics for the pre-European model (left) and the post-European model (right). Boxplots represent distributions calculated on 100 network graphs simulated from the estimated posterior distribution. Red lines represent distributions of observed networks, and gray lines show the 95% intervals."}
library(magick)
pre_bgof_img <- image_read(here::here("analysis/figures/003-pre-bgof.png"))
post_bgof_img <- image_read(here::here("analysis/figures/004-post-bgof.png"))
image_append(c(pre_bgof_img, post_bgof_img))
```

<!--assessment of MCMC analysis and BGOF--> 
Bayesian goodness of fit (GOF) diagnostics plots (Figure \@ref(fig:figure-bgof-two-phases)) show that both models fit the observed networks very well for the minimum geodesic distance distribution (i.e. the number of edges between node pairs in a shortest path [@Hunter2008]) and the degree distribution. For edgewise shared partner (gwesp) distribution, despite some observations falling outside the 95% interval, the fit is generally good with most observations within it. This also suggests that our 
models can reproduce networks that resemble the structural features of our observed networks. We compared the first three distribution moments of each observed distribution and their corresponding simulated distributions, represented by means. Figure \@ref(fig:figure-distribution-moments) shows that modeled values from the pre-European network are slightly closer to the observed values for the mean and the variance of distributions, compared to the values from the post-European network. In general, this demonstrates a better fit for the pre-European network. However, it should be noted that the differences between the statistics of two models are relatively small, which still allows direct comparison of two networks.

```{r figure-distribution-moments, fig.cap = "Distribution moments (mean, skewness, variance) calculated on the observed data and simulated distributions for pre- and post- European models."}
knitr::include_graphics(here::here("analysis", "figures", "006-distribution-moments.png"))
```

@@

<!--archaeological implication based on the results-->
The covariants for both networks, before and after the European presence, provide further information to give insights into how specific elements of the archaeological record relate to the formation of networks (Table \@ref(tab:xxx)). Our covariants were age, sex, ritual pottery, and a wealth index based on the value of grave goods. The pre-European network shows that burials with pottery as ritual vessels tend to be tied. On the other hand, in the post-European network ties tend to be between burials within the same level of wealth. Moreover, in the post-European network, burials with ritual vessels or same sex tend not to be tied. This illustrates that different social and economic mechanisms determined relations between burials for the two periods. Before European arrival, ritual behaviors seem to have been important status indicators, relative to the use of prestige goods. After European presence was established, status indicators shifted from ritual to wealth differences and sex is not associated with status. Physical distance between burials did not correlate to network formation for either period. This means geographical proximity, implying some kinship association, is not a factor for sharing similar goods between burials. This may suggest that the use of burial goods is more individual-oriented instead of kinship-based. In addition, demographic information, such as age and sex, do not show any correlations with network formation for the pre-European network and only sex shows small effect for the post European network. This may be due to the unavailability of age and sex values for many burials in our sample.
